---
title: "Capstone project: HWISE WASH data from Bangladesh"
author: "Rebecca Laes-Kushner"
format: html
editor: visual
source: charity:water
date: today
output: 
  html_document:
  toc: true
  toc_depth: 2
code_fold: true
#bibliography: ref.bib
---

# Introduction

This project analyses baseline data from a water, sanitation and hygiene (WASH) project in Bangladesh run by charity:water. The survey asks about access to water, water issues, handwashing practices and defecation (pooping) practices.

I analyzed responses by categories, such as town (upzila), to see if certain groups have more barriers (water not clean, water further away) and better/worse WASH practices.

# Methods

## Libraries

```{r}
install.packages("tidyverse")
install.packages('here')
install.packages('readxl')
install.packages('gt')
install.packages('dplyr')
install.packages('knitr')
library(here)
library(tidyverse)
library(readxl)
library(gt)
library(dplyr)
library(knitr)
```

## Data import

```{r}
raw <- read_excel(
  here::here("data/raw/Final Baseline and_HwiseSatisfaction to upload raw.xlsx"),
  na = c("", "NA", "N/A", "."))

raw <- raw %>%
  mutate(
    across(where(is.integer), ~ replace(.x, is.na(.x), 0L)),
    across(where(is.double),  ~ replace(.x, is.na(.x), 0))
  )


```

## Prepare data

```{r}

processed <- raw |>

#anonymize towns
mutate(upazila = case_when(
    upazila == "Koyra"      ~ "Upazila A",
    upazila == "Morrelganj" ~ "Upazila B",
    TRUE                    ~ "Upazila C"
))|>
  
#avg time to get water - all
  mutate(avg_time_get_water= round(mean(total_time_to_get_water, na.rm = TRUE)),3) |>
    relocate(avg_time_get_water, .after=total_time_to_get_water) |>

#total water per household, per capita
  mutate(total_water=(((num_25L_filled*25)+(num_20L_filled*20)+(num_15L_filled*15)+(num_10L_filled*10)+(num_5L_filled*5)))) |>
      relocate(total_water, .after=other_containers) |>
    mutate(water_per_cap=round(total_water/num_hh_members,2))|>
      relocate(water_per_cap, .after=total_water)
 
#break out water issues by town
avg_by_upazila <- processed |>
    group_by(upazila) |>
  summarize(
    avg_time_upazila = round(mean(total_time_to_get_water, na.rm = TRUE), 1),
    avg_water_upazila = round(mean(water_per_cap, na.rm=TRUE),1),
    median_water_upazila = round(median(water_per_cap, na.rm=TRUE),1),
    min_water_upazila = min(water_per_cap, na.rm=TRUE),
    max_water_upazila = round(max(water_per_cap, na.rm=TRUE),1)
  )

#summarize when wash hands by gender
cols <- c("when_wash_hands_after_poop","when_wash_hands_after_clean_baby","when_wash_hands_before_cook","when_wash_hands_before_eat",         "when_wash_hands_before_feed_child","when_wash_hands_before_pray","when_wash_hands_after_working_in_dirt")

wash_by_gender <- processed %>%
  group_by(gender) %>%
  summarise(
    n = n(),
      across(all_of(cols), ~ sum(.x, na.rm = TRUE),
                  .names = "sum_{.col}") %>%
  mutate(
  across(starts_with("sum_"), ~ round(.x / n, 1), .names = "pct_{.col}")
) %>%
  rename_with(~ sub("^pct_sum_", "pct_", .), starts_with("pct_")) %>%
  ungroup()
)


```

# Results

Table @tbl-avg-by-upazila2 shows that time spent collecting water daily, upazila C spends about 26% less time (42 minutes) than upazilas A and B, who both spend about 57 minutes. Interestingly, residents in upazila C also collect less water on average per day - almost half of what residents collect in upazila A on average. The median amount of water that Upazila C residents collect is approximately twice as much what residents in the other two upazilas collect daily, with a maximum value that is roughly five times upazila A's maximum value and over twice the maximum value of upazila B.

```{r}
#table by upazila

avg_by_upazila |>
  gt() |>
  fmt_number(columns=c(avg_time_upazila,avg_water_upazila,median_water_upazila,min_water_upazila,max_water_upazila)) |>
  fmt_number(columns = avg_time_upazila:max_water_upazila, decimals = 1) |>
  
cols_label(
  avg_time_upazila = "avg time",
  avg_water_upazila = "avg water",
  median_water_upazila = "median water",
  min_water_upazila = "minimum water",
  max_water_upazila = "maximum water"
) |>
  
tab_header(
  title=md("By upazila, time to collect water and amount of water (l) collected daily")
  )

#| label: tbl-avg-by-upazila2
#| tbl-cap: "average by upazila"
avg_by_upazila |>
    kable(digits = 1)


```

# Conclusions

# References
