---
title: "index.qmd"
format: html
editor: visual
---

## Project goals

Project would be more interesting comparing baseline and endline data...

This data is endline data about WASH issues - how much water people use, source of water, handwashing practices, defecation practices, water availability, etc. I will look at responses by categories, such as demographics, membership in a WASH group and regional district, to see if certain groups have more barriers (water not clean, water further away) and better/worse WASH practices.

```{r}
library(tidyverse)
library(readxl)
```

```{r}
read_xlsx(here::here("data/raw/Final Baseline and_HwiseSatisfaction to upload.xlsx"))
```

```{r}
library(tidyverse)
library(here)

raw_data <- read_xlsx(here::here("data/raw/Final Baseline and_HwiseSatisfaction to upload.xlsx"))

#will rename remaining columns in Excel

processed_data <- raw_data # |>
#  select(column1, column2, column3) |>
#  filter(!is.na(column1)) |>
#  rename(new_name = old_name)

write_csv(processed_data,
          here::here("data/processed/baseline.csv"))


```
